---
title: "Demographic Analysis with DemogAnr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demographic Analysis with DemogAnr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4, fig.align = "center"
)
options(width = 90)
library(DemogAnr)
```

## Why this package

Applied demographic work is still carried out largely in spreadsheets: a life
table here, a total fertility rate there, an age distribution split into single
years. Spreadsheets are visual and familiar, but their formulas are invisible
until each cell is inspected, and reproducing an analysis a year later often
means rebuilding it from memory.

**DemogAnr** aims to make the standard textbook methods directly usable in R.
Three commitments shape it:

* **Textbook notation.** Rates are requested by their conventional
  abbreviations (`"CBR"`, `"TFR"`, `"CDR"`), and life-table columns are named as
  they appear in Preston, Heuveline and Guillot (2001): `nMx`, `nax`, `nqx`,
  `lx`, `dx`, `Lx`, `Tx`, `ex`.
* **Readable output.** Functions return a classed object whose `print` method
  lays the numbers out as a labelled table, so you see what you would write in a
  report rather than a nested list.
* **A consistent API.** Every function takes a data frame plus character column
  names.

Wherever possible, the examples below reproduce a published worked example, so
that you can follow the calculation in the textbook alongside the code.

## Data bundled with the package

| Data set | Contents |
|---|---|
| `gphc2010` | Population and deaths by age, 2010 Ghana Population and Housing Census |
| `ghmort2021` | Population and deaths by age and sex, Ghana 2021 |
| `ghpop2021` | Single-year age distribution by sex, Ghana 2021 |
| `region2000` | Regional base-year parameters, Ghana 2000 |
| `standards` | Brass African and General-UN standard life tables |
| `data` | A grouped (five-year) age distribution |

## 1. Rates: fertility and mortality

`dem.fert()` computes the crude birth rate, general fertility rate,
age-specific rates and the total fertility rate. Here is the Bangladesh 1974
schedule from United Nations *Manual X*, Chapter II, whose reported total
fertility rate is 4.83.

```{r fert}
bangladesh <- data.frame(
  age         = c("15-19","20-24","25-29","30-34","35-39","40-44","45-49"),
  women       = c(3014706, 2653155, 2607009, 2015663, 1771680, 1479505, 1135129),
  live_births = c(320406, 609269, 561494, 367833, 237297, 95357, 38125),
  population  = 71315944
)
fert <- dem.fert(bangladesh, type = "all", age_col = "age",
                 population_col = "population", women_col = "women",
                 births_col = "live_births", graph = FALSE)
fert$results$TFR
```

`dem.cdr()` does the same for mortality. Preston et al. (2001, Box 1.2) report a
crude death rate of 11.47 per 1000 for Sweden in 1988:

```{r cdr}
sweden88 <- data.frame(population = 8438477, deaths = 96756)
dem.cdr(sweden88, type = "CDR", population_col = "population",
        deaths_col = "deaths", graph = FALSE)$results$CDR
```

Childhood and maternal mortality are available through `dm.chm()` and
`dem.mmr()`.

## 2. Life tables

`lifetable()` builds a complete or abridged life table from death rates, or from
deaths and population. Interval widths are taken from the age vector, and the
separation factors ${}_na_x$ follow Coale–Demeny by default.

```{r lt}
lt <- lifetable(gphc2010, age = "Age", pop = "Pop", Dx = "Deaths",
                graph = FALSE)
head(lt$lifetable)
lt$metrics$LifeExpectancyAtBirth
```

Passing `nax_method = "keyfitz"` refines the adult separation factors
iteratively; passing a `nax` vector borrows them from any external model life
table. `lifetable_nqx()` builds a table directly from a series of ${}_nq_x$.

## 3. Multiple decrement and cause-deleted tables

When deaths are classified by cause, `multiple_decrement()` distributes the
life-table deaths across causes. This reproduces Preston et al. (2001) Box 4.1
for US females in 1991: of 100,000 newborns, 21,205 (21.2%) would eventually die
of neoplasms.

```{r mdlt}
us91 <- data.frame(
  Age  = c(0, 1, seq(5, 85, 5)),
  Dall = c(15758, 3169, 1634, 1573, 3955, 4948, 6491, 9428, 12027, 15543,
           19264, 25384, 37211, 59431, 88087, 114693, 143554, 164986, 320578),
  Dneo = c(63, 275, 268, 217, 318, 467, 856, 1924, 3532, 5958,
           8434, 11673, 17078, 25263, 33534, 36695, 36571, 30220, 32739),
  lx   = c(100000, 99217, 99050, 98959, 98870, 98637, 98379, 98070, 97653,
           97083, 96289, 95008, 93018, 89882, 85249, 78711, 69618, 57486, 41756),
  nqx  = c(0.00783, 0.00168, 0.00092, 0.00090, 0.00236, 0.00262, 0.00314,
           0.00425, 0.00584, 0.00818, 0.01330, 0.02095, 0.03371, 0.05155,
           0.07669, 0.11552, 0.17427, 0.27363, 1.00000),
  nax  = c(0.152, 1.605, 2.275, 2.843, 2.657, 2.547, 2.550, 2.616, 2.677,
           2.685, 2.681, 2.655, 2.647, 2.646, 2.631, 2.628, 2.618, 2.570, 6.539)
)
us91$Dother <- us91$Dall - us91$Dneo

md <- multiple_decrement(us91, age = "Age", causes = c("Dneo", "Dother"),
                         deaths = "Dall", lx = "lx", nqx = "nqx", graph = FALSE)
round(md$cause_distribution, 3)
```

`cause_deleted_lt()` answers the companion question: what would life expectancy
be if the cause were eliminated? The deletion assumption is yours to choose
through `method`.

```{r cdlt}
cd <- cause_deleted_lt(us91, age = "Age", cause = "Dneo", deaths = "Dall",
                       lx = "lx", nqx = "nqx", nax_all = "nax", graph = FALSE)
c(e0 = cd$e0, without_neoplasms = cd$e0_deleted, gain = cd$gain)
```

Both functions accept either a population column, from which the life table is
built, or a ready-made table through `lx`, `nqx` and `nax_all`.

## 4. Reproduction and the stable population

`reproduction()` computes the gross and net reproduction rates. For US females
in 1991 (Preston et al. 2001, Box 5.5) the published values are GRR 1.013 and
NRR 0.995.

```{r repro}
us1991 <- data.frame(
  age   = seq(10, 45, 5),
  fbir  = c(5816, 253979, 532712, 596823, 431694, 162005, 25531, 829),
  women = c(8620, 8371, 9419, 10325, 11125, 10344, 9496, 7188) * 1000,
  nLx   = c(494603, 493804, 492552, 491138, 489356, 486941, 483577, 478475)
)
r <- reproduction(us1991, age = "age", births = "fbir", women = "women",
                  nLx = "nLx", radix = 1e5, fraction_female = 1, graph = FALSE)
c(GRR = r$GRR, NRR = r$NRR)
```

`stable_population()` solves Lotka's equation for the intrinsic growth rate by
Coale's iterative procedure. Preston et al. (2001, Box 7.1) obtain
$r = 0.01424$ for Egypt in 1997:

```{r stable}
egypt <- data.frame(
  age = seq(15, 45, 5),
  Lx  = c(4.66740, 4.63097, 4.58518, 4.53206, 4.46912, 4.39135, 4.28969),
  ma  = c(0.00567, 0.06627, 0.11204, 0.07889, 0.05075, 0.01590, 0.00610)
)
sp <- stable_population(egypt, age = "age", asfr = "ma", nLx = "Lx",
                        radix = 1, fraction_female = 1, graph = FALSE)
c(r = sp$r, b = sp$b, d = sp$d, NRR = sp$NRR)
```

## 5. Indirect estimation

Where registration is incomplete, fertility and child mortality are estimated
indirectly. `pf_ratio()` implements the Brass P/F ratio method; on the *Manual
X* Bangladesh 1974 example the adjustment factor is 1.500 and the adjusted total
fertility rate 7.24.

```{r pf}
bd <- data.frame(
  age    = seq(15, 45, 5),
  women  = c(3014706, 2653155, 2607009, 2015663, 1771680, 1479505, 1135129),
  ceb    = c(1160919, 4901382, 9085852, 9910256, 10384001, 9164329, 6905673),
  births = c(320406, 609269, 561494, 367833, 237297, 95357, 38125)
)
pf <- pf_ratio(bd, age = "age", women = "women", ceb = "ceb",
               births = "births", graph = FALSE)
c(K = pf$K, TFR = pf$TFR, TFR_adjusted = pf$TFR_adjusted)
```

`chm_brass()` estimates child mortality from children ever born and children
surviving, by the Brass-Trussell method. It returns $q(x)$ with the calendar
date each estimate refers to.

```{r chm}
panama <- data.frame(
  women = c(2695, 2095, 1828, 1605, 1362, 1128, 930),
  ceb   = c(557, 2633, 4757, 6085, 6722, 6367, 5276),
  cd    = c(40, 130, 312, 435, 636, 686, 689)
)
est <- chm_brass(panama, women_col = "women", ceb_col = "ceb",
                 cd_col = "cd", model = "west", survey_year = 1976.7)
est$qx[est$x == 5]
```

On its own that is a set of probabilities, not a life table. `model_lifetable()`
finishes the job: give it the estimated $q(5)$ and it returns the model table
matching it, in any of the four Coale-Demeny families or the five United
Nations patterns.

```{r mlt}
fitted <- model_lifetable("west", sex = "male", q5 = est$qx[est$x == 5],
                          graph = FALSE)
fitted$metrics$LifeExpectancyAtBirth
```

The same child mortality implies different adult mortality in different
families, which is why the family has to be chosen rather than assumed:

```{r mlt-families}
sapply(c("CD_West", "CD_North", "CD_South", "CD_East"), function(f)
  model_lifetable(f, sex = "male", q5 = est$qx[est$x == 5],
                  graph = FALSE)$metrics$LifeExpectancyAtBirth)
```

The alternative to picking a family is the Brass relational logit model, which
fits a two-parameter transformation of a standard rather than choosing from a
grid. `brass_logit()` estimates the level $\alpha$ and the age-pattern slope
$\beta$; `complete = TRUE` carries them through to the life table they imply.

```{r brass}
fit <- brass_logit(data.frame(age = est$x, qx = est$qx),
                   qx_col = "qx", age_col = "age", standard = "African",
                   complete = TRUE)
round(fit$coefficients, 4)
fit$lifetable$metrics$LifeExpectancyAtBirth
```

`brass_lifetable()` can also be used on its own, with no data at all: give it
$\alpha$ and $\beta$ directly, or a target level to solve for.

```{r brass-target}
brass_lifetable(target_e0 = 60, standard = "African",
                graph = FALSE)$brass$alpha
```

Adult mortality is estimated the same way round. Where `chm_brass()` uses a
mother's report on her children, `orphanhood()` uses children's reports on
their mother, and returns female survivorship from age 25 with the number of
years before the survey each estimate refers to. On the *Manual X* Bolivia
1975 example:

```{r orphanhood}
bolivia <- data.frame(
  age   = seq(15, 50, 5),
  alive = c(5540, 3995, 2886, 1910, 1272, 855, 541, 234),
  dead  = c(448, 541, 769, 852, 945, 1059, 985, 788),
  s10   = c(NA, 0.9060, 0.8401, 0.7474, 0.6313, 0.5175, 0.3996, NA)
)
orph <- orphanhood(bolivia, age = "age", alive = "alive", dead = "dead",
                   prop10 = "s10", M = 28.8, survey_year = 1975.5,
                   graph = FALSE)
orph$table[, c("age", "W", "lx_ratio", "t", "ref_date")]
```

Because each age group of respondents looks a different distance into the
past, one survey yields a series of estimates spread over roughly 8 to 15
years, which is what makes the method useful for a trend rather than a level.

## 6. Completeness of death registration

A life table built from registered deaths is only as good as the registration.
Where it is incomplete -- the normal case wherever civil registration is still
developing -- the table understates mortality, and the first task is to find
out by how much. The death distribution methods do that from two censuses and
the deaths recorded between them.

```{r ddm}
d <- data.frame(
  age = c(0, seq(5, 85, 5)),
  n1  = c(41255, 36289, 31948, 28121, 24743, 21758, 19116, 16770, 14677,
          12798, 11094, 9527, 8060, 6659, 5298, 3970, 2706, 2614),
  n2  = c(52973, 46596, 41022, 36108, 31770, 27938, 24545, 21533, 18846,
          16433, 14245, 12233, 10349, 8550, 6802, 5098, 3475, 3356),
  dth = c(516, 134, 126, 123, 126, 136, 154, 184, 229, 294, 385, 508,
          668, 865, 1084, 1282, 1380, 2833)
)
g <- ggb(d, age = "age", pop1 = "n1", pop2 = "n2", deaths = "dth",
         t1 = 2000, t2 = 2010, graph = FALSE)
c(completeness = g$completeness, coverage_ratio = g$coverage_ratio)
```

These deaths were registered at 70 per cent completeness, and the generalized
growth balance recovers it. `seg()` estimates the same quantity by synthetic
extinct generations, and `ggb_seg()` runs the two in sequence, using the first
to correct for the censuses covering the population differently before the
second is applied.

```{r ddm-compare}
c(GGB = g$completeness,
  SEG = seg(d, "age", "n1", "n2", "dth", 2000, 2010, graph = FALSE)$completeness,
  combined = ggb_seg(d, "age", "n1", "n2", "dth", 2000, 2010,
                     graph = FALSE)$completeness)
```

Neither method should be used without looking at its diagnostic plot first:
`plot()` on a `ggb()` result shows the fitted line, which should be straight,
and on a `seg()` result the age-specific ratios, which should be flat. A curve
in either is a sign that the assumptions -- a closed population, registration
equally complete at every age -- do not hold.

## 7. Standardization and decomposition

Crude rates are not comparable across populations with different age structures.
`standardize()` compares two populations against a common standard, and
`decompose_rates()` splits the difference into a rate component and an
age-composition component.

```{r std}
male   <- subset(ghmort2021, Sex == "Male")
female <- subset(ghmort2021, Sex == "Female")
comp <- data.frame(
  age  = male$Age,
  m_m  = male$Deaths / male$Population,
  m_f  = female$Deaths / female$Population,
  n_m  = male$Population,
  n_f  = female$Population
)
standardize(comp, "age", rate1_col = "m_m", rate2_col = "m_f",
            pop1_col = "n_m", pop2_col = "n_f",
            labels = c("Males", "Females"))
```

`decompose_LE()` attributes a difference in life expectancy at birth to each age
group, following Arriaga.

## 8. Projection

`math_project()` performs deterministic subnational projection by six methods
(linear, exponential, geometric, constant-share, shift-share, share-of-growth),
optionally constraining units to a parent control total.

`cohort_component()` performs the full age–sex projection through a Leslie
matrix. It reproduces Preston et al. (2001) Box 6.1 for Swedish females exactly:

```{r ccm}
sweden <- data.frame(
  Age = c(seq(0, 80, 5), 85),
  N93 = c(293395, 248369, 240012, 261346, 285209, 314388, 281290, 286923,
          304108, 324946, 247613, 211351, 215140, 221764, 223506, 183654,
          141990, 112424),
  Lx  = c(497487, 497138, 496901, 496531, 495902, 495168, 494213, 492760,
          490447, 486613, 480665, 471786, 457852, 436153, 402775, 350358,
          271512, 291707),
  Fx  = c(0, 0, 0, 0.0120, 0.0908, 0.1499, 0.1125, 0.0441, 0.0074, 0.0003,
          rep(0, 8))
)
cc <- cohort_component(sweden, age = "Age", population = "N93", nLx = "Lx",
                       asfr = "Fx", steps = 2, graph = FALSE)
round(cc$totals)
```

`project_population()` carries uncertainty through the demographic balancing
equation $g = b - d + m$, drawing each component from a distribution that you
specify.

```{r pproj}
pp <- project_population(region2000[1:4, ],
        base_year = 2000, future_year = 2010,
        region_var = "Country", subregion_var = "Region",
        base_pop_var = "base_pop", birth_rate_var = "cbr",
        death_rate_var = "cdr", net_migration_var = "nmr",
        num_samples = 500, graph = FALSE)
subset(pp, year == 2010)
```

Each trajectory draws its rates once and holds them across the horizon, so the
band widens with the projection. Supplying `death_dist`, `birth_dist` or
`migration_dist` replaces the defaults with your own sampler.

`project_population()` can also carry an assumed trend rather than holding
each rate constant. `birth_target`, `death_target` and `migration_target` give
the value a component is heading towards by `future_year` (a single number, or
a column of per-region targets), and `path` chooses whether it gets there
`"linear"`ly or `"exponential"`ly. Each trajectory then draws both a start rate
and an end rate and interpolates between them, so the trend and the
uncertainty about it travel together; `target_cv` sets the spread around the
target and is normally wider than `cv`, since a rate two decades out is less
certain than today's:

```{r pproj-trend}
pp_trend <- project_population(region2000[1:4, ],
        base_year = 2000, future_year = 2030,
        region_var = "Country", subregion_var = "Region",
        base_pop_var = "base_pop", birth_rate_var = "cbr",
        death_rate_var = "cdr", net_migration_var = "nmr",
        birth_target = 0.024, death_target = 0.008,
        cv = 0.05, target_cv = 0.20,
        num_samples = 500, graph = FALSE)
subset(pp_trend, year == 2030)
```

Omitting the targets holds every rate at its base-year value and reproduces
the earlier, constant-rate behaviour exactly.

`calibrate_regions()` addresses a different problem: not projecting each
region, but disaggregating a **national projection that is already settled**
down to regions, raking the regions back onto the national control total in
every year. Where `project_population()` lets each region find its own total,
`calibrate_regions()` takes the national trajectory as given and asks only how
it is shared out, which is the sequence used in official projection work
(first agree the national numbers, then split them regionally):

```{r calreg}
base_total <- sum(region2000$base_pop)
national <- data.frame(
  year  = seq(2000, 2020, 5),
  total = round(base_total * exp(0.025 * seq(0, 20, 5)))
)
cal <- calibrate_regions(region2000,
        region_col = "Region", pop_col = "base_pop",
        national = national, year_col = "year", total_col = "total",
        deviation = 0.08, num_samples = 500, graph = FALSE)
cal$check
```

Each region is allowed to deviate from the national profile by a
mean-preserving lognormal (here with coefficient of variation 0.08), phased in
over the horizon, and then raked back onto the national total; `cal$check`
confirms the regional sum reproduces the control total in every year. With no
`deviation` the function reduces to the ordinary constant-share allocation and
every raking factor is 1.

## 9. Data quality

Reported ages should be screened before analysis. `whipple()` and `myers()`
measure digit preference, and `un_age_sex_accuracy()` combines age and sex
ratios into the United Nations joint index.

```{r quality}
males <- subset(ghpop2021, Sex == "Male")
w <- whipple(males, age_col = "Age", pop_col = "Population", graph = FALSE)
w$index
```

`smam()` computes the singulate mean age at marriage from proportions single by
age. Preston et al. (2001, Box 4.4) report 25.0 years for Turkish males in 1990:

```{r smam}
turkey <- data.frame(
  age    = seq(15, 50, 5),
  men    = c(3165061, 2581153, 2435765, 2096899, 1784121, 1418784,
             1111113, 980115),
  single = c(3030203, 1853222, 629077, 180767, 77134, 43412, 28627, 22527)
)
smam(turkey, age = "age", single = "single", total = "men")$SMAM
```

## 10. Figures

Core functions take `graph = TRUE`, attaching a **ggplot2** figure retrievable
with `plot()`. A standalone `pyramid()` draws a population pyramid.

```{r pyramid, fig.height = 5}
g <- within(ghpop2021, grp <- ifelse(Age >= 80, "80+",
            paste0((Age %/% 5) * 5, "-", (Age %/% 5) * 5 + 4)))
pyr <- aggregate(Population ~ grp + Sex, data = g, FUN = sum)
pyramid(pyr, age_col = "grp", sex_col = "Sex", count_col = "Population",
        title = "Ghana, 2021")
```

## References

Coale, A. J., & Trussell, T. J. (1974). Model fertility schedules.
*Population Index*, 40(2), 185–258.

Hajnal, J. (1953). Age at marriage and proportions marrying.
*Population Studies*, 7(2), 111–136.

Preston, S. H., Heuveline, P., & Guillot, M. (2001). *Demography: Measuring and
Modeling Population Processes*. Oxford: Blackwell.

Siegel, J. S., & Swanson, D. A. (Eds.) (2004). *The Methods and Materials of
Demography* (2nd ed.). San Diego: Elsevier.

United Nations (1983). *Manual X: Indirect Techniques for Demographic
Estimation*. New York: United Nations.
